---
title: "betti_numbers(): counting holes"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

Once a set of relations has been made into a
[simplicial complex](build_simplicial.html), it has a shape — and a shape can
be counted. Betti numbers count the holes, one number per dimension:

- **b0** — connected components. How many separate pieces?
- **b1** — one-dimensional holes. Loops that are *not* filled in.
- **b2** — two-dimensional voids. Cavities enclosed by triangles.

The useful ones are `b1` and `b2`, because they detect **absence**. A loop in
`b1` is a cycle of relations where every consecutive pair is present but the
interior is not — a near-miss. Frequency-based summaries cannot see those;
they only report what happened.

```{r load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
```

## Counting

```{r betti}
betti_numbers(sc)
```

One connected component, two independent loops, no voids. The two loops are
the finding: pairs of nodes are connected all the way round a cycle, but the
cycle is never completed into filled triangles.

## The Euler characteristic

`euler_characteristic()` compresses the whole shape into one integer — the
alternating sum of face counts.

```{r euler}
euler_characteristic(sc)
```

It can be computed two entirely different ways, and they must agree. That is
the **Euler-Poincaré identity**, and it is the best available check that the
homology computation is correct.

```{r identity}
fv <- as.data.frame(sc, what = "f_vector")
fv

# from the faces: f0 - f1 + f2 - f3
sum(fv$count * (-1)^fv$dim)

# from the holes: b0 - b1 + b2 - b3
b <- betti_numbers(sc)
sum(b * (-1)^(seq_along(b) - 1))
```

All three give −1. If they diverge, something is wrong — this identity is
worth asserting in any pipeline that computes homology.

## Betti numbers are threshold-dependent

This is the caveat that matters most.

```{r sweep}
do.call(rbind, lapply(seq(0.3, 0.75, by = 0.05), function(t) {
  s <- build_simplicial(m, type = "clique", threshold = t)
  b <- betti_numbers(s)
  data.frame(threshold = t, simplices = s$n_simplices,
             b0 = b[["b0"]], b1 = b[["b1"]],
             euler = euler_characteristic(s))
}))
```

`b1` is **0, then 2, then 4, then 0** across a modest range of thresholds. A
paper reporting "the complex has two loops" without the threshold is
reporting an artefact of an arbitrary choice.

Two honest ways to handle it:

1. **Show the sweep**, as above, and describe how the count behaves.
2. Use **[`persistent_homology()`](persistent_homology.html)**, which sweeps
   the threshold automatically and reports how *long* each feature survives.
   A loop that persists across many scales is real; one that appears at a
   single threshold is noise.

## Seeing the sweep

The table above is a curve. Plotting it is the honest way to report a
threshold-dependent invariant: it shows the whole range rather than the one
threshold that happened to be chosen.

```{r sweepplot, fig.width = 8, fig.height = 5, fig.alt = "Betti numbers b0 and b1 across the threshold sweep"}
long <- do.call(rbind, lapply(seq(0.3, 0.75, by = 0.05), function(t) {
  b <- betti_numbers(build_simplicial(m, type = "clique", threshold = t))
  data.frame(threshold = t, dimension = c("b0", "b1"),
             betti = c(b[["b0"]], b[["b1"]]))
}))
ggplot2::ggplot(long, ggplot2::aes(x = threshold, y = betti,
                                   colour = dimension, shape = dimension)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 2.2) +
  ggplot2::scale_colour_manual(values = c(b0 = "#0072B2", b1 = "#D55E00")) +
  ggplot2::labs(x = "threshold", y = "Betti number", colour = NULL,
                shape = NULL,
                title = "Betti numbers across the threshold sweep") +
  ggplot2::theme_minimal(base_size = 12)
```

## On real data

```{r real}
seqs <- split(human_long$code, human_long$session_id)
hon <- build_hon(seqs, max_order = 3, min_freq = 5)
sc_hon <- build_simplicial(hon, type = "pathway")
sc_hon
betti_numbers(sc_hon)
```

`b0 = 1`: the pathway complex is connected — no group of actions is isolated.
`b1 = 0`: every loop of actions is filled in by some observed pathway.
`b2 = 18`: there are **18 independent two-dimensional voids** — configurations
of four actions where every triple was observed but the four-way combination
never was.

Those 18 are the corpus's near-misses: patterns the sessions came right up to
the edge of and never completed. That is a claim about what is *absent*,
which is why it needed a topological method to find.

## What to do next

- **[`persistent_homology()`](persistent_homology.html)** — the threshold
  problem, solved properly.
- **[`q_analysis()`](q_analysis.html)** — a different notion of
  connectivity: how strongly do the groups overlap?
- **[`build_simplicial()`](build_simplicial.html)** — the constructions these
  numbers describe.

## References

- Edelsbrunner, H., & Harer, J. (2010). *Computational Topology: An
  Introduction*. American Mathematical Society.
- Ghrist, R. (2008). Barcodes: The persistent topology of data. *Bulletin of
  the American Mathematical Society*, 45(1), 61-75.
