---
title: "build_simplicial(): from a network to a shape"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

A graph records which *pairs* of things are related. A simplicial complex
records which *groups* are related — and, crucially, it is **closed under
subsets**: if `{A, B, C}` is a relation, so are `{A, B}`, `{A, C}` and
`{B, C}`.

That closure is not bookkeeping. It is what gives the object a geometry, and
therefore a topology: connected pieces, loops, voids. Those are properties no
pairwise summary can express.

`build_simplicial()` builds one three ways, selected with `type =`.

```{r load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

## `type = "clique"` — from a network

Every clique in the thresholded graph becomes a simplex. A triangle of
mutually-connected nodes becomes a filled 2-simplex.

```{r clique}
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
sc
```

Ten nodes carry 51 simplices up to dimension 3: 24 edges, 15 triangles and 2
tetrahedra.

```{r table}
as.data.frame(sc, top = 5)
as.data.frame(sc, what = "f_vector")
```

The **f-vector** counts faces by dimension and is the complex's basic
fingerprint. `dim = 0` are vertices, `1` edges, `2` triangles, `3`
tetrahedra.

```{r bydim}
as.data.frame(sc, dim = 3, top = 3)
```

## Seeing the complex

`plot()` draws the face counts by dimension and the complex itself.

```{r scplot, fig.width = 13, fig.height = 6, fig.alt = "Face counts by dimension for the clique complex"}
plot(sc)
```

## `threshold` decides the shape

This is the consequential parameter, and its effect is not monotone in any
simple way.

```{r sweep}
do.call(rbind, lapply(c(0.3, 0.4, 0.5, 0.6, 0.7), function(t) {
  s <- build_simplicial(m, type = "clique", threshold = t)
  b <- betti_numbers(s)
  data.frame(threshold = t, simplices = s$n_simplices, dimension = s$dimension,
             euler = euler_characteristic(s),
             b0 = b[["b0"]], b1 = b[["b1"]])
}))
```

Watch `b1`, the number of one-dimensional holes. It is **0 at threshold 0.3,
rises to 4 at 0.6, then collapses to 0 at 0.7**. At a low threshold the graph
is dense enough that every loop is filled in; at a high one it is too sparse
to form loops at all. The interesting structure lives in between.

That instability is the whole argument for
[`persistent_homology()`](persistent_homology.html), which sweeps the
threshold instead of committing to one. **Never report a Betti number from a
single arbitrary threshold** without showing the sweep.

## `type = "vr"` — from a distance matrix

The Vietoris-Rips complex connects points closer than a scale, then fills in
every resulting clique. It is the standard construction for point-cloud data.

```{r vr}
set.seed(2)
pts <- matrix(rnorm(40), 20, 2)
d <- as.matrix(dist(pts))
sv <- build_simplicial(d, type = "vr", threshold = 1.5, max_dim = 3L)
sv
```

### `max_dim` is not optional here

A VR complex grows combinatorially. On this 20-point cloud:

```{r vrcost}
do.call(rbind, lapply(c(2L, 3L, 4L), function(md) {
  t0 <- proc.time()[["elapsed"]]
  s <- build_simplicial(d, type = "vr", threshold = 1.5, max_dim = md)
  data.frame(max_dim = md, simplices = s$n_simplices,
             seconds = round(proc.time()[["elapsed"]] - t0, 2))
}))
```

Roughly a 3-4× increase per dimension. At the **default `max_dim = 10L`**
this same 20-point cloud produces **784,625 simplices** and takes minutes —
so for VR complexes, set `max_dim` deliberately. Dimensions above 3 rarely
carry interpretable structure, and `persistent_homology()` defaults to
`max_dim = 3L` for exactly this reason.

## `type = "pathway"` — from a memory network

The third route takes a fitted model from the memory family and turns its
extracted paths into simplices. See
[`pathways()`](pathways.html) for what each model contributes.

```{r pathway}
seqs <- split(human_long$code, human_long$session_id)
hon <- build_hon(seqs, max_order = 2, min_freq = 50)
build_simplicial(hon, type = "pathway")
```

Here `b0 = 2` says the well-evidenced pathways split into two disconnected
groups of actions — a statement about the corpus that the transition network
cannot make.

## Checking the construction

`verify_simplicial()` re-derives the f-vector independently, as a guard
against a mis-specified threshold or a malformed matrix.

```{r verify}
verify_simplicial(m, threshold = 0.5)
```

It also confirms the **Euler-Poincaré identity** — the alternating sum of the
f-vector equals the alternating sum of the Betti numbers:

```{r euler}
fv <- as.data.frame(sc, what = "f_vector")
sum(fv$count * (-1)^fv$dim)
euler_characteristic(sc)
```

Both give −1. If they ever disagree, the homology computation is wrong.

## What to do next

- **[`betti_numbers()`](betti_numbers.html)** — what the holes mean, and how
  to read them.
- **[`persistent_homology()`](persistent_homology.html)** — the answer to the
  threshold problem above.
- **[`q_analysis()`](q_analysis.html)** — connectivity at each level of
  overlap, and per-node higher-order degree.
- **[`build_hypergraph()`](build_hypergraph.html)** — the same cliques
  without the subset-closure requirement.

## References

- Edelsbrunner, H., & Harer, J. (2010). *Computational Topology: An
  Introduction*. American Mathematical Society.
- Zomorodian, A., & Carlsson, G. (2005). Computing persistent homology.
  *Discrete & Computational Geometry*, 33(2), 249-274.
