---
title: "build_hypergraph(): promoting cliques to group interactions"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

Sometimes only pairwise data is available, but the process being studied is
plausibly a group process. Three people who all know each other might form a
genuine three-way relationship — or they might just be three separate
friendships that happen to close a triangle. A graph cannot tell you which.

`build_hypergraph()` takes the position of Burgio et al. (2020): treat each
triangle (and larger clique) as a **candidate** group interaction, and
promote it to a hyperedge with probability `p`. At `p = 1` every clique
becomes a group; at `p = 0` none do and the pairwise structure is preserved
unchanged. The parameter is an explicit statement of how much you believe
triadic closure implies group interaction.

If you *observe* group membership directly, do not infer it — use
[`group_hypergraph()`](group_hypergraph.html) instead. If your data are
sequences, use [`window_hypergraph()`](window_hypergraph.html).

```{r load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## Building one

```{r build}
hg <- build_hypergraph(m, threshold = 0.5)
hg
```

24 pairwise edges retained as 2-hyperedges, plus 15 triangles promoted to
3-hyperedges.

```{r table}
as.data.frame(hg, top = 6)
```

`weight` is `NA` here: hyperedges from cliques are unweighted. Only
[`window_hypergraph()`](window_hypergraph.html) produces weights, from window
counts.

## Seeing it

A hypergraph has no direct network drawing, but `clique_expansion()`
projects it back to a pairwise network that cograph plots directly. Read it
knowing what the projection destroys: every hyperedge became a clique, and
the group boundaries are gone.

```{r hgplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the hypergraph"}
splot(clique_expansion(hg), minimum = 0.05,
      title = "Hypergraph, projected to a pairwise network")
```

## `p` is the modelling assumption

```{r p}
do.call(rbind, lapply(c(0, 0.25, 0.5, 0.75, 1), function(p) {
  h <- build_hypergraph(m, threshold = 0.5, p = p, seed = 1)
  data.frame(p = p, hyperedges = h$n_hyperedges,
             pairwise = sum(lengths(h$hyperedges) == 2),
             triples = sum(lengths(h$hyperedges) == 3))
}))
```

At `p = 0` you get 24 hyperedges, all of size 2 — the original graph,
re-expressed. At `p = 1`, all 15 triangles are promoted. In between, the
sampling is Bernoulli per clique, so **`seed` is required for
reproducibility**:

```{r seed}
identical(build_hypergraph(m, threshold = 0.5, p = 0.5, seed = 42),
          build_hypergraph(m, threshold = 0.5, p = 0.5, seed = 42))
```

Never report a result at `0 < p < 1` from a single draw. Repeat across seeds
and report the variability, exactly as you would for any stochastic method.

```{r seedvar}
sizes <- vapply(1:20, function(s) {
  build_hypergraph(m, threshold = 0.5, p = 0.5, seed = s)$n_hyperedges
}, integer(1))
range(sizes)
```

## `include_pairwise` — a hypergraph of groups only

```{r nopairs}
build_hypergraph(m, threshold = 0.5, include_pairwise = FALSE)
```

`FALSE` drops the 2-hyperedges, giving a "fully higher-order" hypergraph
containing only the k ≥ 3 interactions. Use it to isolate the group structure
from the dyadic structure — for instance to check whether a centrality result
survives when the pairwise scaffolding is removed.

## `max_size` and `threshold`

```{r maxsize}
do.call(rbind, lapply(c(3, 4, 5), function(ms) {
  h <- build_hypergraph(m, threshold = 0.5, max_size = ms)
  data.frame(max_size = ms, hyperedges = h$n_hyperedges)
}))
```

`max_size = 3` (the default) keeps triangles. Raising it to 4 finds two
4-cliques; there are no 5-cliques, so 5 adds nothing. Unlike the Vietoris-Rips
complex in [`build_simplicial()`](build_simplicial.html), this does **not**
explode combinatorially, because only maximal-clique membership is
enumerated rather than every face.

`threshold` binarises the weight matrix before clique-finding, and it moves
the result far more than `max_size` does:

```{r threshold}
do.call(rbind, lapply(c(0.3, 0.5, 0.7), function(t) {
  h <- build_hypergraph(m, threshold = t)
  data.frame(threshold = t, hyperedges = h$n_hyperedges,
             pairwise = sum(lengths(h$hyperedges) == 2),
             triples = sum(lengths(h$hyperedges) == 3))
}))
```

## `type = "vr"` raises rather than pretending

```{r vr, error = TRUE}
build_hypergraph(m, type = "vr")
```

A genuine metric Vietoris-Rips construction needs a distance-ball graph,
which this verb does not build. Rather than silently aliasing `"clique"` and
giving you a result labelled something it is not, it stops. Use
[`build_simplicial(type = "vr")`](build_simplicial.html) if you want a real
VR construction.

## What to do next

- **[`hypergraph_measures()`](hypergraph_measures.html)** — hyperdegree,
  overlap and density for the object you just built.
- **[`hypergraph_centrality()`](hypergraph_centrality.html)** — which nodes
  matter, by tensor eigenvector or EDVW PageRank.
- **[`clique_expansion()`](clique_expansion.html)** — project back to a
  pairwise network, and see what the round trip costs.
- **[`group_hypergraph()`](group_hypergraph.html)** — when membership is
  observed rather than inferred.

## References

- Burgio, G., Matamalas, J. T., Gómez, S., & Arenas, A. (2020). Evolution of
  cooperation in the presence of higher-order interactions: from networks to
  hypergraphs. *Entropy*, 22(7), 744.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.
