---
title: "build_hypa(): which paths are anomalous"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

A frequent path is not necessarily a surprising one. `Specify -> Command` is
common largely because `Specify` and `Command` are both common — the count
tells you about the states, not about the pairing.

`build_hypa()` (LaRock et al. 2020) separates the two. It asks: given how
often each state is entered and left, **how often would this path occur by
chance?** The null is a Wallenius non-central hypergeometric model, which
holds the marginal in- and out-degrees fixed and lets the paths fall where
they may. A path that beats its own expectation is carrying information the
degree structure does not explain.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

```{r simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

## Running it

```{r run}
hp <- build_hypa(seqs, order = 2)
hp
```

Of the 702 second-order paths, 139 occur significantly more often than
expected and 46 significantly less, after Benjamini-Hochberg correction.

## Over-represented paths

```{r over}
as.data.frame(hp, what = "over", sort_by = "ratio", top = 6)
```

The top row occurs **10.3 times** more often than the degree structure
predicts. That is the signature of a routine — a sequence the sessions
actually follow, rather than one the marginals happen to produce.

## Under-represented paths are the more interesting half

```{r under}
as.data.frame(hp, what = "under", sort_by = "ratio", top = 4)
```

`Specify -> Specify -> Command` is observed 77 times against an expectation
of 126 — a ratio of 0.61. It is a *frequent* path (77 occurrences is not
rare) that is nonetheless **avoided**. A count-based analysis would list it
among the common transitions and miss entirely that it happens far less than
it should.

Absences are hard to see and easy to under-report. This is the column that
finds them.

## Seeing the anomalies

`plot_simplicial()` dispatches on `net_hypa` and ranks pathways by ratio, so
the most extreme deviations come first. `dismantled = TRUE` gives each path
its own panel, which is what makes overlapping routines readable.

```{r overplot, eval = has_cograph, fig.width = 14, fig.height = 14, fig.alt = "Nine most over-represented paths, one panel each"}
plot_simplicial(hp, anomaly = "over", max_pathways = 9,
                dismantled = TRUE, ncol = 3)
```

The under-represented half — the paths the sessions *avoid* — is the harder
one to see in a table and the easier one to see here:

```{r underplot, eval = has_cograph, fig.width = 14, fig.height = 7, fig.alt = "Six most under-represented paths, one panel each"}
plot_simplicial(hp, anomaly = "under", max_pathways = 6,
                dismantled = TRUE, ncol = 3)
```

## Ratio and p-value rank differently

```{r ranks}
as.data.frame(hp, what = "over", sort_by = "ratio", top = 3)
as.data.frame(hp, what = "over", sort_by = "observed", top = 3)
```

The most *surprising* paths and the most *frequent* surprising paths are
different sets. Sort by `ratio` to find structure; sort by `observed` to find
structure that carries weight. Report which you used.

## Multiplicity is handled, and directional

702 paths are tested, so raw p-values would be meaningless. `build_hypa()`
applies Benjamini-Hochberg by default and reports **separate adjusted values
for each direction** — `p_adjusted_over` and `p_adjusted_under` — because a
path cannot be simultaneously over- and under-represented and pooling them
would be wrong.

```{r padjust}
as.data.frame(hp, sort_by = "p", top = 4)
```

`p_value` is the two-sided value; the directional adjusted columns are the
ones to act on. `alpha` and `p_adjust` control the threshold and method:

```{r strict}
hp_strict <- build_hypa(seqs, order = 2, alpha = 0.001, p_adjust = "bonferroni")
hp_strict
```

Under Bonferroni at α = 0.001 the anomaly count falls sharply — the paths
that survive are the ones worth naming in a result.

## `min_count` and order

```{r params}
do.call(rbind, lapply(c(5, 20), function(mc) {
  h <- build_hypa(seqs, order = 2, min_count = mc)
  data.frame(min_count = mc, scored = nrow(as.data.frame(h)),
             over = h$n_over, under = h$n_under)
}))
```

Higher orders test longer paths, and the table grows quickly — check that
each path still has enough observations to support a claim.

## What to do next

- **[`build_hon()`](build_hon.html)** — HYPA scores paths against a null;
  BuildHON+ decides which contexts deserve their own node. They answer
  different questions about the same sequences.
- **[`bootstrap_hon()`](hon_inference.html)** — resampling-based stability
  for extracted rules.
- **[`path_dependence()`](path_dependence.html)** — per-context divergence
  rather than per-path anomaly.
- **[`pathways()`](pathways.html)** — pass the scored paths on to the
  simplicial family.

## References

- LaRock, T., Nanumyan, V., Scholtes, I., Casiraghi, G., Eliassi-Rad, T., &
  Schweitzer, F. (2020). HYPA: Efficient detection of path anomalies in time
  series over networks. *Proceedings of the 2020 SIAM International
  Conference on Data Mining*, 460-468.
- Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery rate.
  *Journal of the Royal Statistical Society B*, 57(1), 289-300.
