---
title: "build_mogen(): choosing an order instead of assuming one"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

[`build_hon()`](build_hon.html) decides *per context* whether memory is
warranted. `build_mogen()` asks the question at the level of the whole model:
fit Markov chains of every order from 0 up to `max_order`, score each one,
and report which order the data actually support.

The multi-order model of Scholtes (2017) treats the layers as a *nested
family* — each order's likelihood is computed on the paths it can explain,
with the degrees of freedom that order costs — so the comparison is a
like-for-like model selection rather than a race between separately-fitted
models.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

```{r simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

## Fitting

```{r fit}
mo <- build_mogen(seqs, max_order = 4)
mo
```

The print method flags the minimum of each criterion, and here they
**disagree**: AIC selects order 4, BIC selects order 1. The asterisks make
the disagreement visible rather than hiding it behind a single "optimal
order".

## The selection table

```{r table}
as.data.frame(mo)
```

The `dof` column explains everything. Going from order 3 to order 4 adds
3,102 parameters (`layer_dof`) on 10,792 observations. The log-likelihood
improves substantially — from −17,163 to −11,366 — so AIC, which charges 2
per parameter, is delighted. BIC charges `log(10792)` ≈ 9.3 per parameter and
is appalled: its score rises from 42,402 at order 1 to 81,061 at order 4.

`optimal` marks the order selected under the criterion you asked for
(`criterion = "aic"` by default).

## Seeing the selection

`plot()` draws both information criteria against order, marking the
selected one. The AIC and BIC curves crossing in different places *is* the
disagreement the next section discusses.

```{r icplot, eval = has_cograph, fig.width = 8, fig.height = 5, fig.alt = "AIC and BIC by Markov order, with the selected order marked"}
plot(mo)
```

## Which criterion should you use?

It depends on what the model is for.

- **Describing the corpus** — AIC. You want the order that best captures the
  structure actually present, and over-parameterisation is a cost you are
  willing to pay for fidelity.
- **Claiming the process is order-k** — BIC. It is consistent for model
  identification, and its verdict of order 1 is a real warning that a
  fourth-order chain is not *identifiable* from 425 sessions.
- **Testing whether order k improves on k−1** — neither; use
  [`markov_order_test()`](markov_order_test.html), which calibrates the
  likelihood-ratio statistic against a permutation null rather than an
  information criterion.

```{r bic}
mo_bic <- build_mogen(seqs, max_order = 4, criterion = "bic")
mo_bic$optimal_order
```

Switching the criterion changes the selected order, and it should be reported
alongside the result.

## The fitted transitions

`mogen_transitions()` is the tidy accessor for the fitted layer.

```{r trans}
mogen_transitions(mo, top = 5)
```

At order 4, a "state" is a four-step history. The heaviest is
`Specify -> Command -> Request -> Specify` continuing to `Interrupt`, 127
times at probability 0.72 — a highly deterministic recurring routine, and the
kind of structure that only a high-order layer can express.

The same table is reachable through the standard accessor, and `order =`
selects a different layer:

```{r layers}
as.data.frame(mo, what = "transitions", order = 2, top = 4)
```

## The fitted layer as shape

`plot_simplicial()` takes a `mogen_transitions()` table directly through
`pathways =`, so the fitted second-order layer can be read as shape over the
network rather than as 702 rows. The network it draws on is the
**first-order** one — nine states. Because `mogen_transitions()` returns
rows already sorted by count, `top = N` picks the N heaviest.

```{r mogensimp, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Six heaviest fitted second-order pathways as simplicial blobs"}
net1 <- build_hon(seqs, max_order = 1)
plot_simplicial(net1, pathways = mogen_transitions(mo, order = 2, top = 6),
                max_pathways = 6,
                title = "Six heaviest second-order pathways")
```

A single pathway, isolated:

```{r mogenone, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The single heaviest fitted pathway"}
plot_simplicial(net1, pathways = mogen_transitions(mo, order = 2, top = 1),
                max_pathways = 1)
```

## Path counts without fitting a model

`path_counts()` is the descriptive counterpart: how often does each k-step
path occur, with no model behind it.

```{r paths}
path_counts(seqs, k = 3, top = 6)
```

Useful as a sanity check before fitting, and for reporting the raw evidence
under a modelled claim.

## A caution on `max_order`

Raising `max_order` is not free. The parameter count grows roughly as
`|S|^(k+1)`, and past a certain point the higher layers are fitting
individual sequences.

```{r cost}
do.call(rbind, lapply(2:4, function(k) {
  m <- build_mogen(seqs, max_order = k)
  d <- as.data.frame(m)
  data.frame(max_order = k, selected_aic = m$optimal_order,
             top_layer_dof = max(d$layer_dof))
}))
```

If AIC keeps selecting whatever ceiling you set, that is a sign the ceiling
is doing the choosing, not the data. Check BIC and the permutation test
before believing it.

## What to do next

- **[`markov_order_test()`](markov_order_test.html)** — the permutation-based
  version of this question, which does not rely on an information criterion.
- **[`build_hon()`](build_hon.html)** — per-context order selection instead
  of one order for the whole model.
- **[`build_hypa()`](build_hypa.html)** — which specific paths occur more or
  less often than a null model expects.
- **[`pathways()`](pathways.html)** — hand the fitted model's paths to the
  simplicial and hypergraph families.

## References

- Scholtes, I. (2017). When is a network a network? Multi-order graphical
  model selection in pathways and temporal networks. *KDD 2017*, 1037-1046.
- Gote, C., & Scholtes, I. (2023). Predicting sequences of traversed nodes in
  graphs using network models with multiple higher orders. *Data Mining and
  Knowledge Discovery*.
