---
title: "build_honem(): embedding a higher-order network"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

Network embeddings put nodes in a vector space so that structurally similar
nodes end up near each other, making the network usable by anything that
takes numeric features — clustering, regression, a downstream classifier.

Applied to a *first-order* network, an embedding can only encode who follows
whom. HONEM (Saebi, Ciampaglia, Kaplan & Chawla 2020) embeds the
**higher-order** network instead, so the vector for `Specify -> Command`
reflects that context's own continuation behaviour, not just `Command`'s.

The method builds higher-order neighbourhood matrices out to `max_power`
steps, sums them into a single deviation matrix, and takes a truncated SVD.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## Embedding

`build_honem()` consumes a fitted network, not raw sequences — build the
higher-order network first and embed that.

```{r build}
hon <- build_hon(seqs, max_order = 3, min_freq = 5)
em <- build_honem(hon, dim = 8)
em
```

233 nodes in 8 dimensions.

## Reading the coordinates

```{r coords}
as.data.frame(em, top = 6)
```

One row per higher-order node. Note what the rows are: `Command` is a
first-order node, `Command -> Command` a second-order context, and
`Command -> Command -> Correct` a third-order one. **All three are embedded
in the same space**, which is what makes the representation useful — a
context and the state it ends in can be compared directly.

## How many dimensions do you need?

```{r variance}
as.data.frame(em, what = "variance")
```

The first dimension carries 24.3% of the squared singular value, the second
16.0%, and the tail flattens quickly — dimensions 4 through 8 each hold
8-10%. There is no sharp elbow, which is honest information: this network has
no small set of dominant structural axes, and an 8-dimensional embedding is
retaining genuinely distributed structure rather than noise.

```{r scree}
em20 <- build_honem(hon, dim = 20)
as.data.frame(em20, what = "variance", top = 6)
cumulative <- cumsum(as.data.frame(em20, what = "variance")$proportion)
data.frame(dims = c(2, 5, 10, 20), cumulative = round(cumulative[c(2, 5, 10, 20)], 3))
```

**`proportion` is relative to the dimensions retained, not to the full
spectrum** — it sums to 1 by construction, so the same axis has a different
share depending on `dim`. At `dim = 20` the first two axes hold 21.9% and the
first ten 63.2%; the 24.3% reported for `dim1` above is its share of an
8-dimensional retention.

Compare like with like when choosing `dim`, and report the value you used.

## `max_power` sets how far the neighbourhood reaches

```{r maxpower}
do.call(rbind, lapply(c(2, 5, 10), function(mp) {
  e <- build_honem(hon, dim = 4, max_power = mp)
  v <- as.data.frame(e, what = "variance")
  data.frame(max_power = mp, dim1_share = round(v$proportion[1], 3),
             top4_share = round(sum(v$proportion), 3))
}))
```

Larger `max_power` accumulates longer-range neighbourhoods. Because every
retained dimension's share is computed within the retained set, these numbers
describe how concentrated the structure is at each reach, not how much total
variance is captured.

## Plotting

```{r plot, fig.width = 7, fig.height = 5}
plot(em, dims = c(1, 2))
```

Read this cautiously. A 2-D projection of an 8-D embedding shows the first
two axes only — and per the variance table, that is 40% of the structure.
Proximity in the plot is suggestive, not conclusive.

## The states the embedding sits over

The embedding has one row per higher-order *node* — 233 of them. That is
far too many to draw: `splot(hon)` is an unreadable hairball, and it is the
wrong picture anyway, because what the coordinates are relative to is the
nine underlying states. Draw those.

```{r honnet, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "First-order network of the nine action codes"}
net1 <- build_hon(human_long, action = "code", actor = "session_id",
                  time = "timestamp", max_order = 1, min_freq = 5)
splot(net1, minimum = 0.05, title = "The nine states behind the embedding")
```

Every point in the scatter above is a *context over these nine states* —
`Command`, `Command -> Command`, `Command -> Command -> Correct` are three
separate rows of the embedding and three separate paths through this graph.

## What an embedding is not

An embedding is a *representation*, not a result. Nothing above is a
statistical claim: there are no p-values, no intervals, and the coordinates
are only identified up to sign and rotation. Two runs on the same input agree
exactly, but a change in `dim` or `max_power` produces a different space, not
a rescaled one.

Use it as input to something that does make claims — and if the question is
"which actions matter", [`hon_centrality()`](hon_centrality.html) answers it
directly and interpretably instead.

## What to do next

- **[`build_hon()`](build_hon.html)** — the network this verb consumes. The
  embedding inherits every choice made there, `min_freq` especially.
- **[`hon_centrality()`](hon_centrality.html)** — an interpretable ranking
  rather than a latent space.
- **[`hypergraph_cluster()`](hypergraph_laplacian.html)** — if the goal is
  grouping rather than coordinates.

## References

- Saebi, M., Ciampaglia, G. L., Kaplan, L. M., & Chawla, N. V. (2020). HONEM:
  Learning embedding for higher order networks. *Big Data*, 8(4), 255-269.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
