---
title: "build_hon(): higher-order networks from sequences"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

A transition network answers "what follows what". It cannot answer "what
follows what, *given how you got here*" — because a first-order node is just
a state, with no memory of the path that reached it.

`build_hon()` builds a network whose nodes carry that memory. A node like
`Specify -> Command` is the state `Command` together with the fact that
`Specify` preceded it, and it keeps its own continuation distribution. The
network only creates such a node when the evidence justifies it: the
BuildHON+ rule of Xu, Wickramarathne & Chawla (2016) adds a higher-order
context **only if** its continuation distribution differs meaningfully from
the lower-order one it would otherwise inherit.

```{r load}
library(honets)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

```{r simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

The data are `human_long`: 10,796 coded human turns from 429 human-AI
pair-programming sessions, nine action codes.

## Building one

Long-format event data goes in directly — name the action, actor and time
columns and `build_hon()` does the splitting.

```{r build}
hon <- build_hon(human_long,
                 action = "code", actor = "session_id", time = "timestamp",
                 max_order = 3, min_freq = 5)
hon
```

Nine states, but **233 nodes**. The extra 224 are memory contexts that
earned their place. `max_order = 3` is a ceiling, not a target: the rule
extraction stops promoting a context as soon as the higher order stops
paying.

`summary()` adds the order breakdown:

```{r summary}
summary(hon)
```

## Reading the rules

`as.data.frame()` is the accessor. Each row is one extracted rule: a source
context, a target state, how often it occurred, and the conditional
probability.

```{r rules}
as.data.frame(hon, sort_by = "count", top = 6)
```

Read the first row as: *of all the times `Specify` was followed by
something, 26.4% of the time it was `Command`* — 740 occurrences. That one
is first-order (`from_order = 1`), which is the baseline the higher-order
rules have to beat.

The genuinely higher-order rules are the ones with `from_order >= 2`:

```{r higher}
as.data.frame(hon, order_min = 2, sort_by = "count", top = 6)
```

## Seeing it

**Do not draw the higher-order network itself.** It has 233 nodes, one per
memory context, and `splot()` on it is an unreadable hairball. The network
you draw is the **first-order** one — nine states — and the higher-order
structure goes on top of it as *pathways*.

```{r net1, eval = has_cograph}
net1 <- build_hon(human_long, action = "code", actor = "session_id",
                  time = "timestamp", max_order = 1, min_freq = 5)
c(nodes = net1$n_nodes, edges = net1$n_edges)
```

`net_hon` is dual-classed `cograph_network`, so cograph draws it with no
conversion step.

```{r splot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "First-order network of the nine action codes"}
splot(net1, minimum = 0.05,
      title = "Human actions in AI-assisted coding")
```

`plot_simplicial()` puts the extracted rules on that same nine-node layout:
each blob is one context and the state it predicts, sources in blue and the
target in red. Repeated states get distinct positions, so
`Command -> Command -> Correct` covers three visible nodes rather than
collapsing onto one. Pass the rules through `pathways =` — this is the same
table `as.data.frame()` returns, read as shape rather than as rows.

```{r simplicial, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Six strongest higher-order rules drawn as simplicial blobs"}
plot_simplicial(net1,
                pathways = as.data.frame(hon, order_min = 2,
                                         sort_by = "count", top = 6),
                max_pathways = 6,
                title = "Six strongest extracted rules")
```

## How much of the network is higher-order?

```{r orders}
table(as.data.frame(hon)$from_order)
```

81 first-order rules — the full 9×9 transition structure, since on data this
dense every pair occurs. Then 441 second-order and 297 third-order rules
were added on top. Roughly nine in ten rules in this network encode memory.

## `min_freq` is the consequential choice

A context is only considered if it occurs at least `min_freq` times. This is
the main lever on network size, and it is a statement about how much
evidence you require before believing a memory effect.

```{r minfreq}
sizes <- do.call(rbind, lapply(c(5, 20, 50, 100), function(mf) {
  h <- build_hon(human_long, action = "code", actor = "session_id",
                 time = "timestamp", max_order = 3, min_freq = mf)
  data.frame(min_freq = mf, nodes = h$n_nodes, edges = h$n_edges,
             max_order_observed = h$max_order_observed)
}))
sizes
```

The network shrinks from 233 nodes to 15 as the evidence bar rises. All four
still observe order-3 contexts — the *strongest* memory effects survive even
a demanding threshold, which is reassuring. Report the threshold with any
result: a claim about a 233-node network and a claim about a 15-node network
are different claims.

## Accepted input

`build_hon()` takes sequences in whichever shape you have them.

```{r inputs}
# a list of character vectors
build_hon(list(c("A", "B", "C", "A", "B"), c("B", "C", "A", "B")),
          max_order = 2, min_freq = 1)

# a wide data.frame, one sequence per row, trailing NAs stripped
build_hon(data.frame(V1 = c("A", "B"), V2 = c("B", "C"), V3 = c("C", NA)),
          max_order = 2, min_freq = 1)
```

It also accepts a fitted `tna` or `netobject` model directly, decoding
integer-coded states back to their labels.

## What to do next

- **[`markov_order_test()`](markov_order_test.html)** — the prior question.
  Is there memory in these sequences at all? Run it before reporting a
  higher-order network as evidence of anything.
- **[`path_dependence()`](path_dependence.html)** — *which* contexts carry
  the memory, one row per context, with the divergence that justifies it.
- **[`bootstrap_hon()`](hon_inference.html)** — do these rules survive
  resampling? A rule extracted once is not yet a finding.
- **[`hon_centrality()`](hon_centrality.html)** — rank the states using the
  higher-order topology.
- **[`build_honem()`](build_honem.html)** — embed the nodes in a vector
  space.
- **[`build_simplicial(type = "pathway")`](build_simplicial.html)** — read
  the same fitted network as a simplicial complex.

## References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
- Saebi, M., Xu, J., Kaplan, L. M., Ribeiro, B., & Chawla, N. V. (2020).
  Efficient modeling of higher-order dependencies in networks: from
  algorithm to application for anomaly detection. *EPJ Data Science*, 9, 15.
