---
title: "clique_expansion(): projecting back, and what it costs"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

Most network tools take a pairwise network. `clique_expansion()` turns a
hypergraph into one, so the whole first-order toolkit becomes available —
plotting, community detection, standard centralities.

It is also the clearest way to *see* what a hypergraph representation is
worth, because the projection is **lossy and irreversible**, and you can
measure exactly how much it throws away.

```{r load}
library(honets)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## Projecting

Every pair of nodes sharing a hyperedge gets an edge, weighted by how much
they share.

```{r project}
hg <- group_hypergraph(human_long, member = "code", group = "session_id")
net <- clique_expansion(hg)
class(net)
c(nodes = net$n_nodes, edges = net$n_edges)
round(net$weights[1:5, 1:5])
```

Nine nodes, 36 edges — the complete graph on nine nodes. Command and
Interrupt share 390 sessions; Frustrate and Inquire share 179.

The return is a `netobject` / `cograph_network`, so it drops straight into
the packages that consume those.

## What `weighted` actually controls

It binarises the **incidence**, not the output. The output is always a count
of shared hyperedges (weighted by incidence values when they exist).

```{r weighted}
dw <- data.frame(member = c("A", "B", "C", "A", "B"),
                 grp    = c("g1", "g1", "g1", "g2", "g2"),
                 n      = c(2, 5, 3, 1, 4), stringsAsFactors = FALSE)
hw <- group_hypergraph(dw, member = "member", group = "grp", weight = "n")
hw$incidence

clique_expansion(hw, weighted = TRUE)$weights
clique_expansion(hw, weighted = FALSE)$weights
```

With `TRUE`, A-B is `2×5 + 1×4 = 14` — the incidence values multiply. With
`FALSE`, A-B is `2` — they simply share two hyperedges.

On a hypergraph whose incidence is already 0/1 the argument does nothing:

```{r noop}
hb <- group_hypergraph(dw, member = "member", group = "grp")
identical(clique_expansion(hb, weighted = TRUE)$weights,
          clique_expansion(hb, weighted = FALSE)$weights)
```

## Seeing the projection

The projected object is a `cograph_network`, so it plots directly — which is
most of the reason to project at all.

```{r projplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The projected pairwise network"}
splot(net, minimum = 0.05, title = "Hypergraph, projected")
```

## The information the projection destroys

This is the part worth internalising. Take a hypergraph with two 3-groups,
project it, and rebuild a hypergraph from the projection:

```{r roundtrip}
d <- data.frame(member = c("A", "B", "C", "A", "B", "D"),
                grp    = c("g1", "g1", "g1", "g2", "g2", "g2"),
                stringsAsFactors = FALSE)
h1 <- group_hypergraph(d, member = "member", group = "grp")
as.data.frame(h1)$states
```

Two genuine three-way groups. Project:

```{r proj}
n1 <- clique_expansion(h1)
n1$weights
```

Rebuild:

```{r rebuild}
h2 <- build_hypergraph(n1$weights, threshold = 0)
as.data.frame(h2)$states
```

**Two hyperedges went in; seven came out.** The two real triples are
recovered — but so are five pairwise edges that were never groups in their
own right, only by-products of the projection. A hypergraph and its
projection are not interchangeable, and the round trip cannot tell a genuine
group from a triangle of coincidences.

That gap is the entire case for
[`build_hypergraph()`](build_hypergraph.html)'s `p` parameter: when you
*must* infer groups from pairwise data, `p` is where you state how much you
believe the inference.

## Projecting collapses the coding sessions completely

```{r collapse}
m <- hypergraph_measures(hg)
as.data.frame(m, what = "global")
```

Pairwise participation is **1.0**: every pair of codes co-occurs somewhere.
So the projection is the complete graph, and any first-order centrality on it
is near-degenerate — the same saturation that makes the first-order
transition network uninformative in the
[three-families comparison](three_families.html).

```{r degenerate}
deg <- rowSums(net$weights > 0)
range(deg)
```

Every node has degree 8. The hypergraph, meanwhile, still distinguishes the
codes clearly by size and overlap structure — see
[`hypergraph_centrality()`](hypergraph_centrality.html).

## When to use it anyway

- To hand a hypergraph to a tool that only accepts pairwise networks.
- To compare a higher-order result against its first-order counterpart, and
  show that the higher-order structure is doing work.
- When Chitra & Raphael's collapse theorem applies: with **edge-independent**
  vertex weights, a random walk on the hypergraph is *equivalent* to one on
  the weighted clique expansion. In that case the projection is lossless for
  walk-based purposes, and the hypergraph adds nothing — which is itself
  worth knowing, and is why
  [`window_hypergraph()`](window_hypergraph.html)'s edge-dependent weights
  matter.

## What to do next

- **[`build_hypergraph()`](build_hypergraph.html)** — the inverse direction,
  with an explicit belief parameter.
- **[`hypergraph_centrality()`](hypergraph_centrality.html)** — centrality
  without projecting.
- **[`hypergraph_measures()`](hypergraph_measures.html)** — the overlap
  structure that projection flattens.

## References

- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172-1181.
- Agarwal, S., Branson, K., & Belongie, S. (2006). Higher order learning with
  graphs. *ICML 2006*, 17-24.
- Tian, Y., & Zafarani, R. (2024). Higher-order network analysis methods.
  *ACM Computing Surveys*.
