---
title: "group_hypergraph(): when membership is observed"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

A great deal of data is *already* group data. Authors on a paper. Students in
a seminar. Codes applied in one session. Items in one basket. The group
membership is recorded directly — nothing needs inferring.

The usual move is to project it to a pairwise network: everyone in a group
gets connected to everyone else. That projection **destroys information**
irreversibly. A five-author paper and ten separate two-author papers can
produce exactly the same graph, and no analysis of the graph can tell them
apart.

`group_hypergraph()` skips the projection. Each group becomes one hyperedge
spanning all its members, and the group size survives into the analysis.

```{r load}
library(honets)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## Building one

Long-format data: one row per (member, group) observation.

```{r build}
hg <- group_hypergraph(human_long, member = "code", group = "session_id")
hg
```

429 sessions become 429 hyperedges over the 9 action codes. The size
distribution is the thing a pairwise projection would have thrown away: **80
sessions used all nine codes**, while five used only one.

```{r table}
as.data.frame(hg, top = 4)
```

## A small worked example

```{r small}
df <- data.frame(
  person  = c("Alice", "Bob", "Carol", "Alice", "Bob", "Dave", "Carol",
              "Dave", "Eve"),
  session = c("S1", "S1", "S1", "S2", "S2", "S3", "S3", "S3", "S3"),
  stringsAsFactors = FALSE
)
small <- group_hypergraph(df, member = "person", group = "session")
small
as.data.frame(small)
```

Three sessions, three hyperedges: a triple, a pair, and another triple. Note
that S1 = {Alice, Bob, Carol} is **one** relation, not three.

## Seeing it

```{r grpplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the group hypergraph"}
splot(clique_expansion(hg), minimum = 0.05,
      title = "Group co-membership, projected")
```

## Weighted membership

If the data record a quantity per (member, group) pair — hours contributed,
messages sent, counts — pass the column as `weight` and the incidence matrix
carries it.

```{r weighted}
dfw <- data.frame(
  person  = c("A", "B", "A", "A", "B"),
  grp     = c("g1", "g1", "g1", "g2", "g2"),
  n       = c(2, 5, 3, 1, 4),
  stringsAsFactors = FALSE
)
hw <- group_hypergraph(dfw, member = "person", group = "grp", weight = "n")
hw$incidence
```

Repeated `(member, group)` pairs are **summed**: A appears twice in `g1` with
2 and 3, giving 5. Without `weight`, the incidence is 0/1 and repeats simply
collapse:

```{r binary}
group_hypergraph(dfw, member = "person", group = "grp")$incidence
```

Those edge-dependent vertex weights are exactly what
[`hypergraph_centrality(type = "pagerank")`](hypergraph_centrality.html) and
the [Laplacian family](hypergraph_laplacian.html) consume — a weighted
hypergraph carries information a binary one does not.

## Missing values are dropped silently

```{r na}
dfn <- data.frame(
  person = c("A", "B", NA, "C", "D"),
  grp    = c("g1", "g1", "g1", NA, "g2"),
  stringsAsFactors = FALSE
)
group_hypergraph(dfn, member = "person", group = "grp")
```

Rows with `NA` in either column are removed before construction. Check your
own missingness before calling — this verb will not warn you about it, and a
silently shrunken hypergraph is easy to miss.

## What the group structure buys you

```{r structure}
m <- hypergraph_measures(hg)
as.data.frame(m, what = "global")
```

**Pairwise participation is 1.0**: every pair of codes co-occurs in at least
one session. A projected pairwise network would therefore be the *complete
graph on nine nodes* — utterly uninformative, exactly as the first-order
transition network is in the
[three-families comparison](three_families.html).

The hypergraph keeps its information because the hyperedges remain
distinguishable by size and membership:

```{r sizes}
table(as.data.frame(m, what = "edges")$size)
```

## Naming

The argument is `member`, not `player`: the values become hypergraph nodes
and need not be people — codes, items, genes and words are all common. The
verb is named for what it returns (a hypergraph) rather than for its input
format.

## What to do next

- **[`hypergraph_measures()`](hypergraph_measures.html)** — hyperdegree,
  overlap structure and density.
- **[`hypergraph_centrality()`](hypergraph_centrality.html)** — which members
  are central across the groups.
- **[`hypergraph_cluster()`](hypergraph_laplacian.html)** — group the members
  by the hypergraph's own random walk.
- **[`clique_expansion()`](clique_expansion.html)** — the projection this
  verb avoids, when you deliberately want it.
- **[`window_hypergraph()`](window_hypergraph.html)** — the sequence-based
  route to a hypergraph, when groups are local in time rather than given.

## References

- Perc, M., Gómez-Gardeñes, J., Szolnoki, A., Floría, L. M., & Moreno, Y.
  (2013). Evolutionary dynamics of group interactions on structured
  populations: a review. *Journal of the Royal Society Interface*, 10(80),
  20120997.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.
