---
title: "Higher-order centrality: ranking actions a first-order network cannot rank"
author: "honets"
date: "2026-08-25"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

Which actions are central to human–AI pair-programming sessions? The
usual instrument is a centrality measure on the transition network: build
the network of what-follows-what, run PageRank, read off the ranking.

That instrument fails on these data, and it fails in an instructive way.
This tutorial shows why, and what a higher-order network recovers.
`hon_centrality()` computes centralities on the higher-order topology —
where a node is not a state but *a state together with the memory of how
it was reached* — and projects them back onto the states, following
Scholtes, Wider & Garas (2016). The implementation follows pathpy 2.2.0,
that paper's reference implementation, generalized to the variable-order
networks `build_hon()` produces.

The data are `human_long`, bundled with honets: 10,796 coded human turns
from 429 sessions, nine action codes.

```{r load}
library(honets)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## The first-order network has nothing to say

```{r firstorder}
fo <- build_hon(human_long, action = "code", actor = "session_id",
                time = "timestamp", max_order = 1, min_freq = 5)
fo
hon_centrality(fo)
```

Every measure returns the same value for all nine actions: PageRank
exactly 1/9, closeness exactly 8, betweenness exactly 0. This is not a
numerical accident. With 10,796 turns spread over nine codes, *every one
of the 81 possible transitions occurs at least five times*, so the
first-order network is the complete digraph on nine nodes. Every node has
in-degree nine, which forces PageRank to be uniform; every pair of states
is one step apart, which forces closeness to be uniform and betweenness
to be zero, because a one-step path has no intermediate node.

The conclusion is decisive and negative: on data this dense, the
first-order network is structurally incapable of ranking the actions. No
choice of centrality measure repairs it — the information is not in the
topology.

## The higher-order network ranks them

```{r higherorder}
hon <- build_hon(human_long, action = "code", actor = "session_id",
                 time = "timestamp", max_order = 3, min_freq = 5)
hon
hon_centrality(hon, type = "pagerank")
```

The same nine actions, now ranked over a 5.6-fold range: Specify holds
23.4% of the flow, Command 18.4%, down to Verify at 4.2%. The topology
that produced this has 233 nodes rather than 9 — the extra 224 are
memory contexts such as `Specify -> Command`, each carrying its own
distinct continuation distribution.

The ranking exists because the *contexts* differ, even though the states
do not. Where the first-order network sees nine interchangeable hubs, the
higher-order network sees which of them recur inside the heavily
travelled contexts. Specify and Command lead because the flow keeps
returning to them along many distinct memory paths; Verify is visited
rarely and from few contexts.

## Seeing the ranking

Two pictures, and the choice between them is the whole point of this
vignette.

First the network that cannot rank. `fo` is the complete digraph on nine
states — every pair one step apart, every node the same degree. The figure
shows why no centrality measure can separate these actions: there is no
structure to separate them by.

```{r foplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "The saturated first-order network: a complete digraph on nine states"}
splot(fo, minimum = 0.05, title = "First-order: saturated, unrankable")
```

Then the ranking itself. Note what is **not** plotted here: the
higher-order network has 233 nodes and drawing it is an unreadable
hairball. `hon_centrality()` already projects onto the nine states, so the
figure is of its tidy return — nine rows, one per action.

```{r centplot, fig.width = 8, fig.height = 5, fig.alt = "PageRank of the nine actions under the higher-order topology"}
pr <- hon_centrality(hon, type = "pagerank", sort_by = "pagerank")
pr$state <- factor(pr$state, levels = rev(pr$state))
ggplot2::ggplot(pr, ggplot2::aes(x = pagerank, y = state)) +
  ggplot2::geom_segment(ggplot2::aes(x = 0, xend = pagerank,
                                     yend = state),
                        colour = "#999999", linewidth = 0.5) +
  ggplot2::geom_point(size = 3.2, colour = "#0072B2") +
  ggplot2::geom_vline(xintercept = 1 / nrow(pr), linetype = "dashed",
                      colour = "#D55E00") +
  ggplot2::annotate("text", x = 1 / nrow(pr), y = 0.7, hjust = -0.1,
                    label = "first-order: all nine sit here",
                    colour = "#D55E00", size = 3.4) +
  ggplot2::labs(x = "PageRank (projected to states)", y = NULL,
                title = "Higher-order topology ranks what first-order cannot") +
  ggplot2::theme_minimal(base_size = 12)
```

The dashed line is 1/9 — where the first-order network puts every action.
The higher-order reading spreads them over a 5.6-fold range.

## Where the ranking comes from

Setting `project = FALSE` reports the centralities of the higher-order
nodes themselves — that is, of the contexts rather than the states.

```{r contexts}
hon_centrality(hon, type = "pagerank", project = FALSE,
               sort_by = "pagerank", top = 8)
```

The single most central object in these sessions is not an action but a
context: `Specify -> Command`, holding 3.5% of the flow on its own — more
than three times any individual first-order node. Below it sit
`Specify -> Specify` and `Command -> Command`, the two self-repetition
contexts, and a third-order context, `Inquire -> Command -> Command`,
already appears in the top eight. Specification followed by commanding is
the load-bearing pattern of this corpus, and it is invisible to any
first-order summary, which can only report that Specify and Command are
both frequent.

## Does the projection choice drive the ranking?

A higher-order node spans several states, so its PageRank has to be
distributed among them. The default, `"scaled"`, splits it equally
(preserving total mass); `"last"` assigns it entirely to the state
actually occupied.

```{r projection}
scaled <- hon_centrality(hon, type = "pagerank", projection = "scaled")
last <- hon_centrality(hon, type = "pagerank", projection = "last")
data.frame(state = scaled$state, scaled = scaled$pagerank,
           last = last$pagerank)
```

The two agree to within about one percentage point on every action, and
the ordering is identical. The ranking is therefore a property of the
higher-order topology, not of the projection rule, and can be reported
without a projection caveat.

## Betweenness and closeness need a sparser network

Both measures were degenerate on the first-order network. They stay
degenerate on the full higher-order network, and for the same reason:

```{r degenerate}
hon_centrality(hon, type = c("betweenness", "closeness"))
```

The projection maps every ordered pair of states to a one-step path,
because a second-order node exists for essentially every pair. With no
intermediate nodes on any shortest path, betweenness is zero everywhere
and closeness is a constant 8. Unlike PageRank, these two measures depend
only on the *reachability* structure, and reachability here is saturated.

Raising `min_freq` keeps only the well-evidenced transitions, which
restores a topology with genuine distances:

```{r sparse}
sparse <- build_hon(human_long, action = "code", actor = "session_id",
                    time = "timestamp", max_order = 3, min_freq = 80)
sparse
hon_centrality(sparse, type = c("betweenness", "closeness"))
```

On the 17-node network of transitions seen at least 80 times, the
measures separate sharply. Specify brokers 22.3 shortest paths and
Command 12.4, while Interrupt, Refine and Verify broker none at all: the
well-evidenced routes between actions run through the two directive
codes, and the remaining actions sit at the ends of those routes rather
than along them. Closeness ranks the same two highest (7.5 and 7.0),
confirming that they are also the easiest actions to reach.

Note what this costs: the sparse network describes only the strongest
transitions, so it answers a narrower question than the PageRank analysis
above. Report the threshold with the result.

## When to use which

- **`hon_centrality(hon)`** — the question is which states carry flow,
  and the sequences have memory. PageRank is the measure to trust on
  dense data; betweenness and closeness require a topology with real
  distances, so check for saturation (all-zero betweenness, constant
  closeness) before interpreting them.
- **`project = FALSE`** — the question is which *contexts* dominate.
  Often the more useful table, and the only place where a pattern like
  `Specify -> Command` can appear at all.
- **`build_hon(max_order = 1)`** — the first-order baseline. Always run
  it: the gap between it and the higher-order ranking is the evidence
  that memory matters, and if there is no gap, the simpler model should
  be reported.
- **[`bootstrap_hon()`](hon_inference.html)** — before reporting a
  higher-order ranking, check that the rules producing it survive
  resampling.
- **`markov_order_test()`** — the prior question of whether memory exists
  at all, as a single global test.

## References

- Scholtes, I., Wider, N., & Garas, A. (2016). Higher-order aggregate
  networks in the analysis of temporal networks: path structures and
  centralities. *The European Physical Journal B*, 89, 61.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5),
  e1600028.
- Scholtes, I. (2017). When is a network a network? Multi-order graphical
  model selection in pathways and temporal networks. *KDD 2017*.
