---
title: "Are the higher-order rules real? Bootstrap and permutation inference for HONs"
author: "honets"
date: "2026-08-25"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 6)
```

## The analytic question

A higher-order network promotes a context to a rule — *after
Specify → Command, the next action follows a different distribution than
after Command alone* — whenever a divergence criterion is met in the
sample (Xu, Wickramarathne & Chawla 2016). The criterion is a point
decision on one dataset. Two questions have to be answered before such
rules are reported as findings:

1. **Stability.** Would the same rules be extracted from another sample
   of the same process? `bootstrap_hon()` resamples sequences with
   replacement and reports, for every rule, the fraction of replicates
   that re-extract it (*support*) and a percentile confidence interval
   for its conditional probability.
2. **Difference.** Do two cohorts follow different rules?
   `compare_hon()` tests every pooled rule edge with a permutation null
   and aggregates a global test, with Benjamini–Hochberg control over the
   per-edge multiplicity.

Both verbs precompute per-sequence counts once and rebuild every
replicate from reweighted counts, so their cost does not grow with
sequence length re-scans, and their results are identical in serial and
parallel runs under the same seed.

The data are `human_long`, bundled with honets: 10,796 coded human turns
from 429 human–AI pair-programming sessions; each session's ordered codes
form one categorical sequence.

```{r load}
library(honets)
```

## Which rules survive resampling?

Rules are extracted up to order 3, requiring each transition to occur at
least 5 times, and every rule is re-examined in 500 bootstrap
resamples of the sessions.

```{r bootstrap}
bs <- bootstrap_hon(human_long, action = "code", actor = "session_id",
                    time = "timestamp", n_boot = 500, max_order = 3,
                    min_freq = 5, seed = 11, parallel = TRUE, n_cores = 2)
bs
summary(bs)
```

The 425 usable sessions yield 829 rule edges: 81 first-order, 451
second-order, 297 third-order. The summary separates the layers sharply.
The second-order layer is solid — mean support 0.995, and even its least
stable rule is re-extracted in 89.8% of replicates. The third-order layer
is mixed: mean support 0.85, minimum 0.35, and its intervals are wide
(mean width 0.70 against 0.20 at order 2). The conclusion is directional:
the second-order structure of these sessions can be reported as a whole;
third-order claims need the per-rule filter below.

```{r stable}
as.data.frame(bs, order_min = 2, min_support = 0.9,
              sort_by = "count", top = 10)
```

604 of the 748 higher-order rules pass a 0.9 support floor; the table
shows the ten most frequent. The strongest patterns are the directive
loop seen from different depths: after `Command -> Request` the next
action is Specify with probability 0.77 (CI 0.72–0.84), and the
third-order context `Specify -> Command -> Request` sharpens this to 0.97
(CI 0.91–1.00, support 0.90). History genuinely narrows the prediction —
the deeper context nearly determines the next action, and the bootstrap
says that narrowing is not a sampling artifact.

```{r bootplot, fig.alt = "Forest plot of rule probabilities with bootstrap confidence intervals"}
plot(bs, top = 15)
```

The forest plot makes the layering visible: first- and second-order rules
carry tight intervals, third-order rules visibly wider ones.

## Did the rules change between the early and late period?

The sessions span 2026-02-05 to 2026-03-08. Splitting the events at the
median date gives two cohorts of comparable size (a session straddling
the cutoff contributes its early turns to the first cohort and its late
turns to the second).

```{r compare}
early <- subset(human_long, session_date < "2026-02-21")
late  <- subset(human_long, session_date >= "2026-02-21")
cmp <- compare_hon(early, late, action = "code", actor = "session_id",
                   time = "timestamp", n_perm = 500, max_order = 3,
                   min_freq = 5, seed = 12, parallel = TRUE, n_cores = 2,
                   names = c("early", "late"))
cmp
summary(cmp)
```

The two results disagree on purpose, and reading them together is the
point of the design. The global test is unambiguous: the
pooled-count-weighted mean absolute difference is 0.0875, and only 1 of
501 permutations reaches it (p = 0.002) — the rule structure did shift
between the periods. Yet not a single one of the 828 edges is
individually significant after Benjamini–Hochberg adjustment. The shift
is *diffuse*: many small, consistent differences spread across the rule
set, none large enough to survive 828-way multiplicity on its own. The
extreme per-edge differences (absolute difference 1.0 at orders 2–3) are
contexts observed in only one period with small counts — exactly the
edges a permutation null neutralizes.

The decisive report is therefore: the interaction rules changed over
time as a whole, and no single rule should be named as the locus of that
change. Naming one anyway — picking the largest difference and reporting
it despite its adjusted p-value — is the multiple-testing error this
verb's two-level design exists to prevent.

```{r compareplot, fig.alt = "Per-edge probability differences between early and late cohorts"}
plot(cmp, top = 15)
```

The difference plot confirms the reading: the largest differences all
plot in the not-significant encoding, and no edge separates from the
pack.

## When to use which

- **`bootstrap_hon()`** — one cohort, and the question is whether the
  extracted rules and their probabilities are stable enough to report.
  Filter with `as.data.frame(x, order_min = 2, min_support = 0.9)` before
  citing individual higher-order rules; report interval widths, not just
  point probabilities.
- **`compare_hon()`** — two cohorts, and the question is whether the rule
  structure differs. Read the global test first; per-edge findings only
  from the BH-adjusted column. A significant global with no significant
  edges is a finding, not a failure: the difference is distributed.
- **`markov_order_test()`** — the prior, coarser question: is there any
  memory beyond order 1 at all, as a single global decision.
- **`path_dependence()`** — a descriptive per-context diagnostic of where
  memory concentrates, without inference; useful for choosing contexts to
  examine before the bootstrap.

## References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5),
  e1600028.
- Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the
  Bootstrap*. Chapman & Hall.
- Good, P. (2005). *Permutation, Parametric and Bootstrap Tests of
  Hypotheses* (3rd ed.). Springer.
- Benjamini, Y., & Hochberg, Y. (1995). Controlling the false discovery
  rate. *JRSS B*, 57(1), 289–300.
