---
title: "hypergraph_measures(): structure without a projection"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

Graph descriptives — degree, density, clustering — all assume relations come
in pairs. Applied to a projected hypergraph they answer questions about the
projection rather than about the data.

`hypergraph_measures()` computes the equivalents directly on the hypergraph:
how many groups is each node in, how large are the groups, and how much do
the groups overlap. The last of these has no graph analogue at all, and is
usually the most informative.

```{r load}
library(honets)
hg <- group_hypergraph(human_long, member = "code", group = "session_id")
m <- hypergraph_measures(hg)
m
```

```{r cograph, eval = has_cograph}
library(cograph)
```

## Per-node measures

```{r nodes}
as.data.frame(m, sort_by = "hyperdegree")
```

- **`hyperdegree`** — how many hyperedges contain this node. Command appears
  in 423 of 429 sessions.
- **`node_strength`** — the summed incidence weight. On a binary incidence
  this is a weighted count; the gap between it and hyperdegree tells you how
  much the weights matter.
- **`max_edge_size`** — the largest group this node ever belonged to. All 9
  here, because every code appears in at least one nine-code session.

The spread is narrow: 423 down to 218, a ratio under 2. Every action is
present in most sessions, so **hyperdegree does not discriminate strongly** —
worth knowing before building an argument on it. Compare with
[`hypergraph_centrality()`](hypergraph_centrality.html), which uses the
overlap structure rather than raw membership counts.

## Whole-hypergraph measures

```{r global}
as.data.frame(m, what = "global")
```

**Density 0.729** is the fraction of possible (node, hyperedge) memberships
that exist — a hypergraph analogue of graph density, and high here because
sessions use most of the code set.

**Pairwise participation 1.0** is the one to notice. It is the fraction of
node *pairs* that co-occur in at least one hyperedge, and it is at ceiling:
every pair of codes shares a session. That is precisely why a pairwise
projection of these data is uninformative — it would be the complete graph.

## Hyperedge sizes

```{r edges}
as.data.frame(m, what = "edges", top = 5)
table(as.data.frame(m, what = "edges")$size)
```

The size distribution is the structure a projection destroys. Here it is
strongly right-skewed: most sessions use six or more of the nine codes.

## Seeing the structure

The per-node table becomes a picture two ways: the co-membership weights as
a heatmap, and the projected network itself. The heatmap is the honest one —
it shows the pairwise weights the measures are computed from without
implying the hypergraph *was* a graph.

```{r heat, eval = has_cograph, fig.width = 8, fig.height = 7, fig.alt = "Heatmap of pairwise co-membership weights"}
plot_heatmap(clique_expansion(hg))
```

```{r hgmplot, eval = has_cograph, fig.width = 9, fig.height = 9, fig.alt = "Clique expansion of the hypergraph"}
splot(clique_expansion(hg), minimum = 0.05,
      title = "Co-membership, projected to a pairwise network")
```

## The intersection profile

The most distinctively hypergraph-shaped output, shown by `print()`: how many
*pairs of hyperedges* share exactly k members.

```{r profile}
m
```

96 pairs of sessions share no codes at all; 20,355 pairs share five. Read as
a distribution, it says these sessions are **highly overlapping** — the
typical pair has five codes in common. A collection of specialised, disjoint
groups would put its mass at the low end instead.

This is a property of the *set system*. No pairwise projection retains it.

## A contrasting example

```{r contrast}
disjoint <- group_hypergraph(
  data.frame(member = c("A","B","C","D","E","F"),
             grp    = c("g1","g1","g2","g2","g3","g3"),
             stringsAsFactors = FALSE),
  member = "member", group = "grp")
md <- hypergraph_measures(disjoint)
as.data.frame(md, what = "global")
md
```

Three disjoint pairs: density 0.33, pairwise participation 0.2, and every
hyperedge pair shares zero members. Compare against the sessions above to see
how far apart two hypergraphs can be on measures that a graph would flatten.

## On an inferred hypergraph

```{r inferred}
set.seed(1)
mat <- matrix(runif(100), 10, 10)
mat <- (mat + t(mat)) / 2
diag(mat) <- 0
dimnames(mat) <- list(LETTERS[1:10], LETTERS[1:10])
mi <- hypergraph_measures(build_hypergraph(mat, threshold = 0.5))
as.data.frame(mi, what = "global")
as.data.frame(mi, sort_by = "hyperdegree", top = 4)
```

Clique-derived hypergraphs have much lower density, because most hyperedges
are size 2 or 3 rather than spanning most of the node set.

## What to do next

- **[`hypergraph_centrality()`](hypergraph_centrality.html)** — ranking
  nodes, where hyperdegree does not separate them.
- **[`hypergraph_cluster()`](hypergraph_laplacian.html)** — grouping nodes
  using the hypergraph's random walk.
- **[`clique_expansion()`](clique_expansion.html)** — the projection, and
  what it costs.
- **[`q_analysis()`](q_analysis.html)** — the overlap question again, in the
  simplicial family, where subset closure is assumed.

## References

- Lee, G., Choe, M., & Shin, K. (2024). A survey on hypergraph
  representation, learning and mining. *Data Mining & Knowledge Discovery*,
  37, 1-39.
- Do, M. T., Yoon, S., Hooi, B., & Shin, K. (2020). Structural patterns and
  generative models of real-world hypergraphs. *KDD 2020*, 176-186.
