---
title: "The hypergraph Laplacian family"
subtitle: "hypergraph_laplacian(), hypergraph_cluster(), hypergraph_transduction()"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

Spectral methods — clustering, label propagation, diffusion — all run off a
Laplacian. Defining one for a hypergraph is not automatic, because a
hyperedge connects many nodes at once and there is no single obvious way to
distribute that connection.

This family implements two answers:

- **`type = "zhou"`** (default) — the weighted normalised Laplacian of Zhou,
  Huang & Schölkopf (2006). Each hyperedge spreads its weight evenly over its
  members.
- **`type = "random_walk"`** — the edge-dependent vertex weight walk of
  Hayashi et al. (2020). A step picks a hyperedge in proportion to its
  weight, then a node within it in proportion to that node's weight *in that
  hyperedge*. When those within-edge weights vary, this carries information
  the Zhou construction averages away.

```{r load}
library(honets)
hg <- window_hypergraph(human_long, action = "code", actor = "session_id",
                        time = "timestamp",
                        window = 3L, min_size = 2L, min_weight = 50L)
hg
```

## The Laplacian itself

```{r lap}
L <- hypergraph_laplacian(hg)
dim(L)
round(L[1:4, 1:4], 3)
```

It is a symmetric 9×9 matrix with the properties a Laplacian should have:

```{r props}
ev <- sort(eigen(L, symmetric = TRUE, only.values = TRUE)$values)
round(head(ev, 4), 5)
```

The smallest eigenvalue is exactly 0 — the constant vector is always in the
kernel. The **second** eigenvalue (0.548) is the spectral gap: how hard it is
to cut this hypergraph in two. A large gap means well-connected, hard to
split.

The stationary distribution of the associated walk comes along as an
attribute, and the [PageRank verb](hypergraph_centrality.html) is the damped
version of the same walk:

```{r pi}
round(sort(attr(L, "pi"), decreasing = TRUE), 4)
```

## Clustering

```{r cluster}
cl <- hypergraph_cluster(hg, k = 3, seed = 1)
cl
```

```{r clustertable}
as.data.frame(cl)
```

Three clusters of three. Reading the membership: **Command, Inquire and
Interrupt** group together; **Correct, Frustrate and Refine** form a second;
**Request, Specify and Verify** a third. The second is recognisably a
repair-and-revision cluster, the third a specify-and-check cluster.

`summary()` gives the sizes:

```{r clsummary}
summary(cl)
```

### Check the eigengap before trusting `k`

The print method reports **Eigengap after k: 0.0380**. That is small. A
well-separated `k`-cluster structure produces a large gap after the `k`-th
eigenvalue; a small one means the choice of `k` is not strongly supported by
the spectrum.

```{r eigengap}
do.call(rbind, lapply(2:5, function(k) {
  c_k <- hypergraph_cluster(hg, k = k, seed = 1)
  data.frame(k = k, sizes = paste(table(as.data.frame(c_k)$cluster),
                                  collapse = "/"))
}))
```

`k` is your assumption, not a discovery. Report the eigengap alongside it, or
show that the conclusion survives several values.

### Seeding

`hypergraph_cluster()` runs k-means on the spectral embedding, which is
stochastic. Pass `seed`:

```{r seed}
identical(hypergraph_cluster(hg, k = 3, seed = 1),
          hypergraph_cluster(hg, k = 3, seed = 1))
```

`nstart` (default 25) controls how many k-means restarts are used; raise it
if clusters move between seeds.

## Seeing the clustering

`plot()` gives two diagnostic panels: the Laplacian spectrum with the
chosen `k` marked, and the spectral embedding the k-means step actually
clustered. The eigengap after `k` is the evidence for (or against) the
choice — read the scree panel before believing the assignment.

```{r clplot, fig.width = 12, fig.height = 6, fig.alt = "Laplacian spectrum and spectral embedding, coloured by cluster"}
plot(cl)
```

## Transductive label spreading

Label a few nodes, and let the hypergraph structure propagate labels to the
rest.

```{r trans}
labs <- stats::setNames(c("directive", "evaluative"), c("Command", "Verify"))
tr <- hypergraph_transduction(hg, labels = labs, xi = 0.5)
tr
as.data.frame(tr)
```

`score` is the confidence in the predicted class; `margin` is the gap to the
runner-up. Small margins mean the structure did not decide strongly.

### The default `xi` can override your own seed labels

`xi` trades off *smoothness over the hypergraph* against *fidelity to the
seeds*. It defaults to 0.99, which weights smoothness heavily — heavily
enough that a seed label can be overturned:

```{r xi}
do.call(rbind, lapply(c(0.1, 0.5, 0.9, 0.99), function(x) {
  d <- as.data.frame(hypergraph_transduction(hg, labels = labs, xi = x))
  data.frame(xi = x,
             Verify_seeded_as = "evaluative",
             Verify_predicted = d$predicted[d$node == "Verify"],
             n_evaluative = sum(d$predicted == "evaluative"))
}))
```

At `xi = 0.99` — **the default** — Verify is predicted `directive` despite
being seeded `evaluative`, and the evaluative class disappears entirely. At
`xi ≤ 0.9` the seed holds.

This is the documented behaviour of Zhou's algorithm, not a bug: with one
seed per class and a strongly-connected structure, 99% smoothing swamps the
1% fidelity term. But it means **you must check that your seeds survive**.
Lower `xi`, or seed more nodes:

```{r moreseeds}
labs2 <- stats::setNames(
  c("directive", "directive", "evaluative", "evaluative"),
  c("Command", "Specify", "Verify", "Correct"))
as.data.frame(hypergraph_transduction(hg, labels = labs2, xi = 0.9))
```

With two seeds per class the evaluative class survives at the same `xi`.

## Seeing the spreading

The full node-by-class score matrix, with seed nodes outlined in black and
each node's winning class marked. Rows where the winning and runner-up
scores are close are the assignments to distrust — that closeness is
visible here and invisible in a table of predicted labels.

```{r trplot, fig.width = 8, fig.height = 6, fig.alt = "Heatmap of transduction scores by node and class"}
plot(hypergraph_transduction(hg, labels = labs2, xi = 0.9))
```

## Both Laplacians

```{r types}
do.call(rbind, lapply(c("zhou", "random_walk"), function(ty) {
  Lt <- hypergraph_laplacian(hg, type = ty)
  e <- sort(eigen(Lt, symmetric = TRUE, only.values = TRUE)$values)
  data.frame(type = ty, lambda_2 = round(e[2], 4),
             lambda_max = round(max(e), 4))
}))
```

Use `random_walk` when the incidence carries genuinely edge-dependent vertex
weights — as `window_hypergraph()` produces, where a cell is an occurrence
count within a window. On a binary incidence the two constructions carry the
same information, and Chitra & Raphael's collapse theorem says the walk is
then equivalent to one on a weighted clique expansion.

## Connectivity is required

```{r disconnected, error = TRUE}
iso <- group_hypergraph(
  data.frame(member = c("A", "B", "C", "D"),
             grp    = c("g1", "g1", "g2", "g2"), stringsAsFactors = FALSE),
  member = "member", group = "grp")
hypergraph_laplacian(iso)
```

A disconnected hypergraph has a degenerate spectrum — the multiplicity of
eigenvalue 0 equals the number of components — so these verbs refuse it with
a classed `honets_hypergraph_disconnected` condition rather than returning
something meaningless. Split the components and analyse them separately.

## What to do next

- **[`hypergraph_centrality()`](hypergraph_centrality.html)** — the damped
  version of this same random walk.
- **[`hypergraph_measures()`](hypergraph_measures.html)** — descriptive
  structure before the spectral analysis.
- **[`window_hypergraph()`](window_hypergraph.html)** — where the
  edge-dependent weights come from.

## References

- Zhou, D., Huang, J., & Schölkopf, B. (2006). Learning with hypergraphs:
  clustering, classification, and embedding. *NeurIPS 19*, 1601-1608.
- Hayashi, K., Aksoy, S. G., Park, C. H., & Park, H. (2020). Hypergraph
  random walks, Laplacians, and clustering. *CIKM 2020*, 495-504.
- Chitra, U., & Raphael, B. J. (2019). Random walks on hypergraphs with
  edge-dependent vertex weights. *ICML 2019*, PMLR 97, 1172-1181.
- von Luxburg, U. (2007). A tutorial on spectral clustering. *Statistics and
  Computing*, 17(4), 395-416.
