---
title: "markov_order_test(): does memory exist at all?"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

Every higher-order method in this package assumes the sequences have memory.
`markov_order_test()` is the question you should ask **before** any of them:
does the past predict the future beyond the present state, and if so, how far
back?

It fits Markov chains of increasing order, compares them by likelihood-ratio,
and — because the asymptotic chi-square reference is unreliable when the
transition table is sparse — calibrates the statistic against a **permutation
null** built by shuffling within sequences.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
wide <- as.data.frame(do.call(rbind, lapply(
  seqs, function(s) c(s, rep(NA, max(lengths(seqs)) - length(s))))),
  stringsAsFactors = FALSE)
```

## Running it

```{r run}
mk <- markov_order_test(wide, max_order = 3, n_perm = 200, seed = 1)
mk
```

Three criteria, and they **disagree**: BIC selects order 1, AIC and the
permutation LRT both select order 3. That disagreement is the result, not a
problem to resolve away.

```{r table}
as.data.frame(mk)
```

## Reading the disagreement

Look at the `df` column. Going from order 2 to order 3 costs **4,721 extra
parameters** on 10,792 observations. AIC charges 2 per parameter; BIC charges
`log(n)` ≈ 9.3. On data this size the two criteria are answering different
questions:

- **AIC** asks which order predicts best out of sample. It says 3.
- **BIC** asks which order is most likely to be the true generating model. It
  says 1, because the parameter cost of a third-order chain is enormous
  relative to the evidence.

Both are right about their own question. For *description* — which is what
`build_hon()` does — the AIC/LRT answer is the relevant one, and it is why
`max_order = 3` is a sensible ceiling for these data. For a claim that the
process **is** third-order Markov, BIC's scepticism should be reported.

## Seeing the test

`plot()` puts the two panels side by side: the information criteria by
order, and the observed statistic against its permutation null. The second
panel is the one to read — it shows *how far* outside the null the observed
value sits, which the p-value's floor cannot express.

```{r mkplot, fig.width = 12, fig.height = 6, fig.alt = "Information criteria by order, and observed statistic against the permutation null"}
plot(mk)
```

## The permutation p-value has a floor

```{r pfloor}
as.data.frame(mk)
```

Every permutation p-value is 0.004975 — that is `1 / (200 + 1)`, the
**smallest value 200 permutations can produce**. It does not mean the three
orders are equally significant; it means all three exceeded every one of
their 200 null replicates. Raise `n_perm` if you need to resolve below that
floor.

Note also how far the asymptotic p-values sit from it: 1.7e-129 at order 2.
That is the sparsity problem the permutation null exists to avoid — with 576
degrees of freedom over nine states, the chi-square approximation is not
trustworthy, and the permutation value is the one to report.

## The null distribution itself

`what = "null"` returns the replicates, so the calibration can be inspected
rather than trusted.

```{r null}
nulls <- as.data.frame(mk, what = "null")
as.data.frame(mk, what = "null", top = 4)
aggregate(g2 ~ order, data = nulls,
          FUN = function(x) round(c(mean = mean(x), max = max(x)), 1))
```

Compare each order's observed `g2` in the table above against the maximum its
null ever reached. The observed statistics are far outside — which is what
"p at the floor" means, made concrete.

## Cost and reproducibility

Permutations dominate the runtime. `parallel = TRUE` uses
`parallel::mclapply` (skipped on Windows), and all randomness is drawn
serially before any parallel work, so **a parallel run reproduces the serial
result exactly under the same seed**.

```{r parallel, eval = FALSE}
markov_order_test(wide, max_order = 3, n_perm = 2000,
                  parallel = TRUE, n_cores = 4, seed = 1)
```

Always pass `seed`. A permutation test reported without one is not
reproducible.

## What to do next

- **[`build_hon()`](build_hon.html)** — if there is memory, build the network
  that represents it. Use the selected order as the `max_order` ceiling.
- **[`path_dependence()`](path_dependence.html)** — this test is global. That
  one localises the memory to individual contexts.
- **[`build_mogen()`](build_mogen.html)** — the same model-selection question
  under a multi-order generative model, which mixes orders rather than
  choosing one.

## References

- Anderson, T. W., & Goodman, L. A. (1957). Statistical inference about
  Markov chains. *The Annals of Mathematical Statistics*, 28(1), 89-110.
- Good, P. (2005). *Permutation, Parametric and Bootstrap Tests of
  Hypotheses*. Springer.
