---
title: "pathways(): the bridge between the three families"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
# cograph is a Suggests dependency. Every plot chunk below carries
# `eval = has_cograph`, so this vignette still knits (without figures) where
# cograph is not installed.
has_cograph <- requireNamespace("cograph", quietly = TRUE)
```

## The analytic question

The three structure families in honets are entry points into one another, not
islands. `pathways()` is the generic that carries a fitted **memory network**
across to the other two: it extracts the paths the model considers real and
returns them as plain strings, which
[`build_simplicial(type = "pathway")`](build_simplicial.html) turns into a
complex.

It is a small verb doing one job, and it is the reason a HON can be handed to
a topological analysis without any manual reshaping.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
```

```{r cograph, eval = has_cograph}
library(cograph)
```

```{r simpguard, include = FALSE, eval = has_cograph}
# grDevices::chull() aborts on a degenerate (collinear or single-point)
# pathway cloud. cograph >= 2.3.5 fixes this in .smooth_blob(); shadowing the
# function here keeps the render alive on CRAN cograph (2.1.1) by falling
# back to splot(). Scope is this vignette's knit environment only.
plot_simplicial <- function(...) {
  args <- list(...)
  tryCatch(do.call(cograph::plot_simplicial, args),
           error = function(e) {
             warning("plot_simplicial fell back to splot(): ",
                     conditionMessage(e), call. = FALSE)
             splot(args[[1]])
           })
}
```

## Three methods, one generic

`pathways()` dispatches on `net_hon`, `net_hypa` and `net_mogen`. Each
returns the paths that model regards as meaningful — which is a *different
set* in each case, because the three models mean different things by
"meaningful".

```{r hon}
hon <- build_hon(seqs, max_order = 2, min_freq = 50)
p_hon <- pathways(hon)
length(p_hon)
pathways(hon, top = 6)
```

From a HON: the extracted rules — contexts that earned their own node. 25 of
them at `min_freq = 50`.

```{r hypa}
hp <- build_hypa(seqs, order = 2)
p_hypa <- pathways(hp)
length(p_hypa)
pathways(hp, top = 4)
```

From HYPA: the **anomalous** paths — those significantly over- or
under-represented against the degree-preserving null. 185 of them. Note the
first entry, `Interrupt Interrupt -> Interrupt`: the same repeated-
interruption pattern that [`path_dependence()`](path_dependence.html) flags
as the highest-divergence context.

```{r mogen}
mo <- build_mogen(seqs, max_order = 2)
p_mogen <- pathways(mo)
length(p_mogen)
pathways(mo, top = 4)
```

From MOGen: every path in the fitted layer — 702, the complete second-order
transition set, because a multi-order model does not prune.

## The format

Each string is `"<context states, space separated> -> <next state>"`. The
context uses spaces so that the arrow separates history from continuation
unambiguously.

```{r format}
p_hon[1]
```

Read as: the context `Specify Command` continuing to `Specify`.

## Crossing to the simplicial family

```{r simplicial}
sc <- build_simplicial(hon, type = "pathway")
sc
```

`build_simplicial()` calls `pathways()` internally — you rarely need to call
it yourself. Each path becomes a simplex, and because a simplicial complex is
closed under subsets, so does every sub-path. The result has a geometry, and
therefore a topology: `b0 = 2` says the well-evidenced pathways fall into
**two disconnected groups** of actions, and `b1 = 1` says there is one
one-dimensional hole — a cycle of actions that is never closed by an observed
pathway.

Neither statement is available from the transition network the paths came
from.

## Seeing them

Whichever model produced them, pathways draw the same way — as blobs over
the **first-order** nine-node layout. The higher-order object supplies the
pathways; it is never the thing drawn. Compare the sources by eye: HON gives
the contexts that earned a node, HYPA the ones that deviate from the null,
MOGen every fitted path.

```{r pwplot, eval = has_cograph, fig.width = 11, fig.height = 11, fig.alt = "Pathways from the HON, drawn as simplicial blobs"}
net1 <- build_hon(seqs, max_order = 1)
plot_simplicial(net1,
                pathways = as.data.frame(hon, order_min = 2,
                                         sort_by = "count", top = 6),
                max_pathways = 6, title = "Pathways from build_hon()")
```

```{r pwhypa, eval = has_cograph, fig.width = 14, fig.height = 7, fig.alt = "Anomalous pathways from HYPA, one panel each"}
plot_simplicial(hp, anomaly = "over", max_pathways = 6,
                dismantled = TRUE, ncol = 3)
```

## The choice of source model matters

Because the three models select different paths, they build different
complexes:

```{r compare}
do.call(rbind, lapply(
  list(HON = hon, HYPA = hp, MOGen = mo),
  function(m) {
    s <- build_simplicial(m, type = "pathway")
    data.frame(paths = length(pathways(m)), simplices = s$n_simplices,
               dimension = s$dimension, b0 = betti_numbers(s)[1])
  }))
```

MOGen's 702 unpruned paths produce a dense, connected complex; HON's 25
well-evidenced rules produce a sparse, disconnected one. **The topology you
report is a property of the model you chose**, not of the raw sequences — say
which model, and say why.

## Calling it directly

You would call `pathways()` yourself to inspect what a model considers real,
or to feed the paths somewhere else:

```{r direct}
# how long are the extracted paths?
table(lengths(strsplit(p_hon, " ", fixed = TRUE)))
```

## What to do next

- **[`build_simplicial()`](build_simplicial.html)** — the main consumer, and
  where the topology gets computed.
- **[`build_hon()`](build_hon.html)**, **[`build_hypa()`](build_hypa.html)**,
  **[`build_mogen()`](build_mogen.html)** — the three producers.
- **[`window_hypergraph()`](window_hypergraph.html)** — the other route out
  of sequence data, into the hypergraph family.

## References

- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions.
  *Physics Reports*, 874, 1-92.
