---
title: "path_dependence(): which contexts carry the memory"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

[`markov_order_test()`](markov_order_test.html) answers a global question:
*is there memory in these sequences?* It returns one verdict for the whole
corpus.

`path_dependence()` localises it. For every order-k context it asks: **how
different is what happens next, once you know the context, from what happens
next if you only know the current state?** The answer is a Kullback-Leibler
divergence in bits, one row per context, so memory stops being a property of
the dataset and becomes a property of specific situations.

```{r load}
library(honets)
seqs <- split(human_long$code, human_long$session_id)
wide <- as.data.frame(do.call(rbind, lapply(
  seqs, function(s) c(s, rep(NA, max(lengths(seqs)) - length(s))))),
  stringsAsFactors = FALSE)
```

## Running it

```{r run}
pd <- path_dependence(wide, order = 2)
pd
```

Two chain-level numbers summarise the corpus: knowing the previous action as
well as the current one buys **0.143 bits** on average, and cuts the
uncertainty about the next action by **0.109 bits**. Those are modest
averages hiding a very uneven distribution.

## The uneven part

```{r top}
as.data.frame(pd, sort_by = "KL", top = 8)
```

`Interrupt -> Interrupt` diverges by **1.065 bits** — seven times the corpus
average. Its entropy drops from 3.02 to 2.03 bits, meaning that once you know
an interruption followed an interruption, you have roughly halved your
uncertainty about what comes next.

## Seeing which contexts carry it

`plot()` is a dot plot of per-context divergence, sized by how often the
context occurs. The long right tail is the point: memory is concentrated in
a handful of situations, not spread evenly.

```{r pdplot, fig.width = 10.5, fig.height = 6, fig.alt = "Per-context KL divergence, sized by context count"}
plot(pd)
```

## `flips` is the column that matters most

`flips` is `TRUE` when the context changes the *modal* prediction — the
single most likely next action differs from what the first-order model would
have said.

```{r flips}
pd_all <- as.data.frame(pd)
table(flips = pd_all$flips)
```

35 of the 81 contexts flip the prediction. That is the practical statement of
what memory buys here: in **43% of situations, a first-order model would name
the wrong most-likely next action**.

```{r flipex}
as.data.frame(pd, sort_by = "KL", top = 5)
```

Read `Interrupt -> Interrupt`: a first-order model predicts `Command` next
(because `Command` is what usually follows `Interrupt`), but conditioned on
the repeat, the modal continuation is another `Interrupt`. Interruptions
cluster, and only the higher-order view sees it.

## KL and entropy drop are different questions

Both columns are reported because they answer different things.

```{r twoviews}
as.data.frame(pd, sort_by = "KL", top = 6)
```

`H_drop` is how much *uncertainty* the context removes. `KL` is how far the
context's distribution has *moved* from the first-order one. A context can
move a long way without becoming more predictable — `Frustrate -> Specify`
has a high KL (0.410) on a small entropy drop (0.154), meaning it reshuffles
which continuations are likely without concentrating them.

## `min_count` guards against noise

Divergence is easy to manufacture from a handful of observations. Contexts
below `min_count` are dropped.

```{r mincount}
do.call(rbind, lapply(c(5, 20, 50), function(mc) {
  p <- path_dependence(wide, order = 2, min_count = mc)
  d <- as.data.frame(p)
  data.frame(min_count = mc, contexts = nrow(d),
             mean_KL = round(mean(d$KL), 3),
             flips = sum(d$flips))
}))
```

The mean KL *falls* as the threshold rises — from 0.186 to 0.157 — which is
evidence that some of the apparent memory at `min_count = 5` is small-sample
noise. The effect is mild here: only two contexts fall below 20 occurrences,
and the flip count holds at 35 until the threshold reaches 50, where 20
sparse contexts drop out and 27 flips remain. That stability is itself
reassuring. Report the threshold regardless.

## Higher orders

```{r order3}
pd3 <- path_dependence(wide, order = 3, min_count = 20)
pd3
```

## What to do next

- **[`markov_order_test()`](markov_order_test.html)** — the global test this
  verb decomposes. Run it first.
- **[`build_hon()`](build_hon.html)** — the contexts with high KL are exactly
  the ones BuildHON+ promotes to higher-order nodes.
- **[`bootstrap_hon()`](hon_inference.html)** — before reporting a specific
  context as a finding, check that it survives resampling.

## References

- Cover, T. M., & Thomas, J. A. (2006). *Elements of Information Theory*
  (2nd ed.). Wiley.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing
  higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
