---
title: "persistent_homology(): which features are real"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

[`betti_numbers()`](betti_numbers.html) counts holes at **one** threshold,
and the count is unstable: change the threshold slightly and loops appear and
vanish. Any single-threshold result is partly an artefact of the choice.

`persistent_homology()` removes the choice. It sweeps the threshold across
the whole range, tracks every feature from the scale where it is **born** to
the scale where it **dies**, and reports how long each survived. Features
that persist across many scales are structural; features that flicker in and
out at one scale are noise.

This turns "how many holes?" — which has no threshold-free answer — into
"which holes are robust?", which does.

```{r load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
```

## Running it

```{r run}
ph <- persistent_homology(m, n_steps = 12, max_dim = 2)
ph
```

## The persistence diagram

One row per feature: the dimension it lives in, where it was born, where it
died, and the difference.

```{r diagram}
as.data.frame(ph, sort_by = "persistence", top = 8)
```

**`persistence` is the column that matters.** The top feature — a
0-dimensional class with persistence 0.93 — is the connected component that
survives the entire filtration. Everything below it is short-lived by
comparison.

Note that `birth` exceeds `death` here. The filtration runs from a high
weight threshold down to a low one, so features are born at high weights and
die as the complex fills in. Persistence is the length of the interval either
way.

## Reading one dimension at a time

```{r dim1}
as.data.frame(ph, dimension = 1, sort_by = "persistence", top = 5)
```

```{r dim0}
as.data.frame(ph, dimension = 0, sort_by = "persistence", top = 5)
```

Dimension 0 is the merge history of connected components — each row is two
pieces joining. Dimension 1 is loops, dimension 2 voids.

## Betti curves

The same information viewed as counts against threshold, rather than as
intervals.

```{r betti}
as.data.frame(ph, what = "betti", top = 8)
```

Read the `b0` column down: **9 components at the highest threshold, then 8,
7, and then 1** — the graph rapidly coalesces into a single piece as the
threshold falls. That transition point is where the network "connects", and
it is visible here without having to pick a threshold in advance.

## Plotting

```{r plot, fig.width = 8, fig.height = 4}
if (requireNamespace("gridExtra", quietly = TRUE)) plot(ph)
```

The diagram plots death against birth; points far from the diagonal are the
persistent features. The Betti curves show the counts across the sweep.

## Choosing `n_steps` and `max_dim`

```{r params}
do.call(rbind, lapply(c(6L, 12L, 24L), function(ns) {
  p <- persistent_homology(m, n_steps = ns, max_dim = 2)
  d <- as.data.frame(p)
  data.frame(n_steps = ns, features = nrow(d),
             max_persistence = round(max(d$persistence), 4))
}))
```

`n_steps` controls the resolution of the sweep — more steps locate births and
deaths more precisely, at linear cost. The **longest-lived feature is stable
across resolutions**, which is the reassurance you want: the headline finding
does not depend on the grid.

`max_dim` is the expensive one. It defaults to `3L` here (rather than the
`10L` that [`build_simplicial()`](build_simplicial.html) uses) precisely
because complexes grow combinatorially with dimension. Raise it only if you
have a reason to expect structure above dimension 3.

## Essential classes

```{r essential}
d <- as.data.frame(ph)
sum(!is.finite(d$death))
```

Features that never die are reported with `death = Inf`. There are none in
this filtration because it runs to completion. When comparing diagrams with
[`bottleneck_distance()`](persistence_landscape.html), essential classes need
explicit handling — the reference implementations differ on whether to
include them.

## Validation

The Z/2 boundary-matrix reduction behind this verb is checked against
**ripser** (via `TDAstats`) in the package's equivalence suite: 30 random
point clouds, maximum discrepancy 4.44e-16 — machine precision.

## What to do next

- **[`persistence_landscape()`](persistence_landscape.html)** — turn a
  diagram into something you can average, plot and compare statistically.
- **[`betti_numbers()`](betti_numbers.html)** — the single-threshold version,
  and why it needs a sweep.
- **[`build_simplicial()`](build_simplicial.html)** — the complexes this
  filters over.

## References

- Edelsbrunner, H., Letscher, D., & Zomorodian, A. (2002). Topological
  persistence and simplification. *Discrete & Computational Geometry*, 28,
  511-533.
- Zomorodian, A., & Carlsson, G. (2005). Computing persistent homology.
  *Discrete & Computational Geometry*, 33(2), 249-274.
- Ghrist, R. (2008). Barcodes: The persistent topology of data. *Bulletin of
  the American Mathematical Society*, 45(1), 61-75.
