---
title: "q_analysis() and simplicial_degree(): connectivity by overlap"
author: "honets"
date: "2026-08-26"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    self_contained: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "")
```

## The analytic question

In a graph, two nodes are connected or they are not. In a
[simplicial complex](build_simplicial.html) there are *degrees* of
connection: two groups might share one member, or three, or six. Q-analysis
(Atkin 1974) takes that seriously.

Two simplices are **q-connected** when they share at least `q + 1` vertices,
possibly through a chain of intermediates. Sweep `q` from high to low and you
get a nested picture: at high `q` only heavily-overlapping groups are linked,
and the structure fragments; as `q` falls, the fragments merge.

Where the merging happens tells you how the system is organised — something
neither a degree distribution nor a Betti number reports.

```{r load}
library(honets)
set.seed(1)
m <- matrix(runif(100), 10, 10)
m <- (m + t(m)) / 2
diag(m) <- 0
dimnames(m) <- list(LETTERS[1:10], LETTERS[1:10])
sc <- build_simplicial(m, type = "clique", threshold = 0.5)
```

## Running it

```{r run}
qa <- q_analysis(sc)
qa
```

## The q-vector

```{r qvec}
as.data.frame(qa)
```

Read it from the top down. At `q = 3` — groups sharing four or more members —
the complex is in **11 separate pieces**. At `q = 2` it is still 10. At
`q = 1` it drops to 2, and at `q = 0` everything is a single component.

The collapse happens between `q = 2` and `q = 1`: this structure is held
together by *pairwise* overlap, not by deep shared membership. Groups here
touch each other lightly and rarely share three or more members. A system
organised around tightly-interlocking teams would show the opposite profile —
few components even at high `q`.

## Seeing the q-vector

```{r qaplot, fig.width = 13, fig.height = 6, fig.alt = "Q-vector: connected components at each q-level"}
plot(qa)
```

## Per-node structure

```{r nodes}
as.data.frame(qa, what = "nodes")
```

`max_q` is the highest level at which each node still participates in a
connected group. Nodes A, C, D, E and I reach `q = 3`; the rest stop at 2.
Those five sit inside the most heavily-overlapping part of the complex.

## Simplicial degree

`simplicial_degree()` is the higher-order analogue of node degree: not "how
many neighbours" but "how many groups of each size am I in?"

```{r degree}
simplicial_degree(sc)
```

`d0` is always 1 — every node is its own vertex. `d1` is ordinary graph
degree. **`d2` and `d3` are where the higher-order information lives**: E is
in 7 triangles and 2 tetrahedra, F in only 2 triangles and no tetrahedra,
though their graph degrees (6 and 4) are not nearly so far apart.

The table is sorted by `total`, so the ranking is immediate.

### Ordinary degree does not predict it

```{r vsdegree}
d <- simplicial_degree(sc)
data.frame(
  measure = c("d1 (graph degree) vs total", "d2 (triangles) vs total"),
  spearman = c(round(cor(d$d1, d$total, method = "spearman"), 3),
               round(cor(d$d2, d$total, method = "spearman"), 3))
)
```

Graph degree correlates with higher-order degree, but imperfectly — the
ranking by `total` is not the ranking by `d1`. A node can have many
neighbours who are not connected to *each other*, which produces edges
without triangles.

### Normalising

```{r norm}
simplicial_degree(sc, normalized = TRUE, top = 4)
```

`normalized = TRUE` divides each count by the maximum possible at that
dimension, so values are comparable across dimensions and across complexes of
different sizes. Use it when comparing datasets; use the raw counts when
describing one.

## On real data

```{r real}
seqs <- split(human_long$code, human_long$session_id)
sc_hon <- build_simplicial(build_hon(seqs, max_order = 3, min_freq = 5),
                           type = "pathway")
as.data.frame(q_analysis(sc_hon))
simplicial_degree(sc_hon)
```

The pathway complex of the coding sessions behaves very differently from the
random network above: it stays in many components down to `q = 2` and only
consolidates at `q = 1`. And in the degree table, `d0` and `d1` are
**identical for every action** — a consequence of the saturated first-order
structure, where every pair of actions co-occurs. All the discrimination is
in `d2` and `d3`, where Specify (32 tetrahedra) separates sharply from Verify
(8).

When `d1` is constant, ordinary degree centrality is uninformative by
construction, and the higher-order degree is the only thing distinguishing
the nodes.

## What to do next

- **[`betti_numbers()`](betti_numbers.html)** — holes rather than overlap;
  the other topological summary of the same complex.
- **[`build_simplicial()`](build_simplicial.html)** — how the complex is
  built, and how sensitive it is to `threshold`.
- **[`hypergraph_measures()`](hypergraph_measures.html)** — overlap
  statistics without the subset-closure assumption.

## References

- Atkin, R. H. (1974). *Mathematical Structure in Human Affairs*. Heinemann.
- Johnson, J. (2013). *Hypernetworks in the Science of Complex Systems*.
  Imperial College Press.
- Estrada, E., & Ross, G. J. (2018). Centralities in simplicial complexes.
  *Journal of Theoretical Biology*, 438, 46-60.
